---
title: "PhosphoEnzymes: human kinases and phosphatases"
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{PhosphoEnzymes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(PhosphoEnzymes)
```

# Overview

`PhosphoEnzymes` provides class-specific master tables for the human
phosphorylation machinery, keyed on base Ensembl gene IDs and typed by substrate
class. Enzymes are typed by what they act on; sequence-family membership is
evidence of lineage, not an override of substrate.

Both classes ship now: kinases via `get_kinases()`, protein phosphatases via
`get_phosphatases()`, and a thin class-agnostic summary spanning both via
`get_phosphoenzymes()`. The non-catalytic regulatory-subunit companion
(`get_phosphatase_regulators()`) is a forthcoming addition.

# Accessing the tables

```{r load-kinases}
k <- get_kinases()
p <- get_phosphatases()
c(kinases = nrow(k), phosphatases = nrow(p))
# substrate is carried as co-equal columns; empty nonprotein_substrate_type = protein-only
table(protein = k$acts_on_protein, nonprotein = k$nonprotein_substrate_type)
```

```{r strict-example, eval = FALSE}
get_kinases(mode = "strict")        # canonical set: >= 1 evidence dimension
get_phosphatases(mode = "strict")
```

The unified summary spans both classes; recover class-specific taxonomy with a
class-routed join back to a master:

```{r unified}
pe <- get_phosphoenzymes()
table(pe$regulator_class, pe$acts_on_protein)
```

# The evidence model

`n_evidence_dimensions` (0-2) counts distinct evidence **classes**:

1. **Structural / evolutionary catalog** -- membership in an expert
   sequence-family catalog (`in_structural_catalog`).
2. **Biochemical** -- a protein-specific 4-digit EC (`is_protein_kinase_ec`,
   EC 2.7.10-2.7.14 for kinases).

Because the catalogs all answer the same fold question from overlapping
scholarship, they roll up into a single axis; GO and the reviewed UniProt keyword
are supplementary support, not counted axes.

`evidence_tier` is a practical prioritization heuristic over those two axes plus
supplementary support:

| tier | condition |
|------|-----------|
| Gold | both axes (structure **and** biochemistry) |
| Silver | one axis **and** supplementary support (experimental GO or reviewed keyword) |
| Bronze | one axis, no supplementary support |
| Provisional | zero axes |

Because Gold requires the biochemical axis, it is common among well-characterized
kinases but will be rarer among phosphatases, whose protein-specific EC coverage
is far thinner -- a real feature of the underlying data, not a scoring artifact.

```{r tier-table}
table(k$evidence_tier)
```

# Session info

```{r session-info}
sessionInfo()
```
