---
title: "PhosphoEnzymes: A Reproducible, Substrate-Aware Reference of Human Kinases and Phosphatases"
author: "Mahmoud Eladawi"
output:
  BiocStyle::html_document:
    toc_float: true
vignette: >
  %\VignetteIndexEntry{PhosphoEnzymes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(PhosphoEnzymes)
```

# Overview

`PhosphoEnzymes` is a reproducible reference of the human phosphorylation
machinery: protein kinases and protein phosphatases, assembled from several
expert catalogs, keyed on base (unversioned) Ensembl gene IDs, and typed by
**what each enzyme acts on**.

The organizing idea is *substrate awareness*. Every gene is placed on two
**independent** axes:

- **Rigor** -- how well the gene's catalytic identity is established (how many
  distinct kinds of evidence support it), summarized as an `evidence_tier`.
- **Substrate** -- the chemical class the enzyme phosphorylates or
  dephosphorylates (protein, lipid, nucleotide, carbohydrate, ...).

Crucially, **sequence-family membership is evidence of lineage, not an override
of substrate**. A gene that sits inside a classical protein-kinase fold is not
forced to be typed as a protein kinase if its biochemistry says it acts on a
lipid. This is the *PI3K / PTEN principle*: PI3K-family members carry a
protein-kinase-like fold but phosphorylate lipids, and PTEN dephosphorylates
*both* a protein substrate and the lipid PIP3. The tables preserve that duality
rather than collapsing each gene to a single label.

Three accessors expose the data, plus the term sets that drove the typing:

- `get_kinases()` -- the human kinase master.
- `get_phosphatases()` -- the human protein-phosphatase master.
- `get_phosphoenzymes()` -- a thin, class-agnostic summary spanning both.
- `get_term_set()` / `validate_term_set()` -- the EC/GO rule tables shipped as
  data, and a linter for custom rule tables.

# Accessing the tables

```{r load-tables}
k <- get_kinases()
p <- get_phosphatases()
c(kinases = nrow(k), phosphatases = nrow(p))
```

Substrate is carried as co-equal boolean columns. `acts_on_protein` and
`acts_on_nonprotein` are not mutually exclusive -- a *dual* enzyme is `TRUE` for
both -- and `nonprotein_substrate_type` is a pipe-delimited list of the
non-protein chemical classes (empty means protein-only):

```{r substrate-cols}
table(
  acts_on_protein    = k$acts_on_protein,
  acts_on_nonprotein = k$acts_on_nonprotein
)
```

The unified summary spans both classes, one row per gene, with
`regulator_class` distinguishing them. It carries only the shared
substrate/evidence columns; recover class-specific taxonomy with a class-routed
join (shown below).

```{r unified}
pe <- get_phosphoenzymes()
table(pe$regulator_class)
```

# Rigor and substrate are orthogonal

The two axes are deliberately decoupled. `evidence_tier` answers *"how sure are
we this gene is a catalytically real kinase/phosphatase?"* It is **substrate
blind**: it never asks what the enzyme acts on. The substrate columns answer a
separate question entirely.

`n_evidence_dimensions` (0--2) counts distinct evidence **classes**:

1. **Structural / evolutionary catalog** -- membership in an expert
   sequence-family catalog (`in_structural_catalog`).
2. **Biochemical** -- a class-specific four-digit EC number
   (protein kinase EC 2.7.10--2.7.14, or protein phosphatase EC).

These two axes map to a practical tier:

| tier        | condition                                              |
|-------------|--------------------------------------------------------|
| Gold        | both axes (structure **and** biochemistry)             |
| Silver      | one axis **and** supplementary GO or reviewed keyword  |
| Bronze      | one axis, no supplementary support                     |
| Provisional | zero axes                                              |

`curated_core` (used by `mode = "strict"`) is simply
`n_evidence_dimensions >= 1`.

Two kinds of *supplementary support* can lift a one-axis gene from Bronze to
Silver. One is an experimental-evidence GO annotation for the catalytic
activity. The other is a **reviewed UniProt keyword** -- a controlled-vocabulary
term that UniProt's expert curators attach to a manually reviewed
(Swiss-Prot) entry; here it is the family marker a curator assigns to the enzyme
(`KW-0418`, *Kinase*, for kinases; `KW-0904`, *Protein phosphatase*, for
phosphatases). Supplementary support is deliberately *not* one of the two rigor
axes -- it only breaks the Bronze/Silver tie within a single axis, never
manufacturing a second one.

Because rigor and substrate are orthogonal, the cross-product is populated in
every cell that the biology allows: there are **Gold-tier enzymes that act on
lipids**, not only on proteins. The tabulation below makes the independence
concrete -- read it as "tier (rows) does not determine substrate (columns)":

```{r orthogonal}
with(p, table(evidence_tier = evidence_tier, substrate = substrate_call))
```

# A dual enzyme, end to end: PTEN

PTEN is the canonical worked example of why the two axes must stay separate. It
is a **Gold-tier** phosphatase (both evidence axes present) that acts on
**both** a protein substrate and a lipid (PIP3). A pipeline that let fold or
family override substrate would collapse PTEN to a single label and lose half
of its biology. Here it is in one row, straight from the shipped table:

```{r pten-trace}
pten <- subset(
  get_phosphatases(),
  symbol == "PTEN",
  select = c(
    symbol, evidence_tier, n_evidence_dimensions,
    acts_on_protein, acts_on_nonprotein,
    substrate_call, nonprotein_substrate_type
  )
)
as.data.frame(pten)
```

`acts_on_protein` is `TRUE`, `acts_on_nonprotein` is `TRUE`, `substrate_call` is
`dual`, and `nonprotein_substrate_type` records the `lipid` substrate -- all
while `evidence_tier` is `Gold`. Tier (rigor) and substrate are reported
independently, exactly as intended.

Every substrate call is auditable. Three provenance columns record *how* the
call was reached -- the evidence that fired, the deciding source, and whether
the available evidence agreed:

```{r pten-provenance}
as.data.frame(subset(
  get_phosphatases(),
  symbol == "PTEN",
  select = c(substrate_evidence, substrate_decider, substrate_concordance)
))
```

# Two knobs: `mode` and `substrate`

`get_kinases()` and `get_phosphatases()` take two orthogonal filters that mirror
the two axes:

- `mode` filters **rigor**: `"comprehensive"` (default; the full membership
  union) or `"strict"` (`curated_core`, i.e. at least one evidence dimension).
- `substrate` filters **what the enzyme acts on**: `"any"` (default),
  `"protein"`, or `"nonprotein"`. A dual enzyme satisfies both `"protein"` and
  `"nonprotein"`.

The two knobs never couple. `mode = "strict"` is **not** a protein filter: a
strictly-curated lipid kinase is kept unless you *also* ask for
`substrate = "protein"`. Any cell of the cross-product is reachable:

```{r two-knob}
nrow(get_kinases()) # comprehensive, any
nrow(get_kinases(mode = "strict")) # >= 1 evidence dim
nrow(get_kinases(mode = "strict", substrate = "protein")) # strict + protein
```

# Term sets as data

The EC and GO rules that drive substrate typing are not buried in code -- they
ship as four declarative CSVs (kinase/phosphatase x EC/GO), readable with
`get_term_set()`:

```{r term-sets}
kinase_go <- get_term_set("kinase", "go")
head(kinase_go[, c("term_id", "substrate", "substrate_subtype", "role")])
```

Each row is a cited rule: a term id, the substrate class it implies, and whether
it contributes to rigor, to substrate typing, or both. Before feeding a
hand-edited rule table back in, lint it with `validate_term_set()`, which
returns one row per issue (empty when clean):

```{r validate}
issues <- validate_term_set() # validates the four shipped defaults
nrow(issues) # 0 -> the shipped set is clean
```

## Re-typing under a custom term set

`get_kinases()` / `get_phosphatases()` accept a `term_sets =` argument: a named
list of EC/GO rule tables (any subset; missing tables fall back to the shipped
defaults). When supplied, the substrate, rigor, and provenance columns are
**recomputed from each gene's raw per-gene evidence** before the `mode` /
`substrate` filters apply -- so you can re-type the catalog under your own rules
**without rebuilding the package**.

This recompute is **GMT-free**: it does not need the (unshipped) gene-set file.
It works against a per-gene evidence sidecar that ships in the package
(`substrate_evidence.csv`), intersecting each gene's recorded GO terms and EC
codes against the term-set accessions you supply.

Passing the *default* term set reproduces the shipped table exactly -- a
round-trip identity that is the contract anchoring every custom recompute:

```{r round-trip}
default_ts <- list(
  kinase_ec = get_term_set("kinase", "ec"),
  kinase_go = get_term_set("kinase", "go")
)
recomputed <- get_kinases(term_sets = default_ts)
identical(
  as.data.frame(recomputed)[, "substrate_call"],
  as.data.frame(get_kinases())[, "substrate_call"]
)
```

To explore an alternative typing, edit a copy of a term set and pass it back.
For example, dropping the lipid-kinase GO rules would stop typing those genes as
lipid kinases (their rigor is untouched -- again, the two axes are independent):

```{r custom-typing}
ts2 <- get_term_set("kinase", "go")
ts2 <- ts2[ts2$substrate_subtype != "lipid" | is.na(ts2$substrate_subtype), ]
retyped <- get_kinases(term_sets = list(kinase_go = ts2))
c(
  shipped_lipid = sum(grepl("lipid", get_kinases()$nonprotein_substrate_type)),
  retyped_lipid = sum(grepl("lipid", retyped$nonprotein_substrate_type))
)
```

The default accessor path uses only base R. The `term_sets =` recompute path is
the *only* code that needs `dplyr`, `purrr`, `stringr`, and `readr`; these are
`Suggests`, not hard dependencies, and are reached solely when you pass a custom
term set.

# Recovering class taxonomy from the unified summary

`get_phosphoenzymes()` keeps only the shared columns, so class-specific taxonomy
(kinase group/family, phosphatase fold/family) lives on the masters. Recover it
with a **class-routed join**: split the summary by `regulator_class` and join
each part back to the matching master on `ensembl_gene_id`.

```{r class-routed-join}
pe <- get_phosphoenzymes()

kin_tax <- merge(
  pe[pe$regulator_class == "kinase", c("ensembl_gene_id", "evidence_tier")],
  get_kinases()[, c("ensembl_gene_id", "kinase_family")],
  by = "ensembl_gene_id"
)
head(kin_tax, 3)
```

# Session info

```{r session-info}
sessionInfo()
```
